---
title: "week2ex"
author: "Nicholas Metaxas"
date: "Saturday, October 01, 2016"
output: html_document
---

##Assignment for Week 2 coursera course Reproducible Research 

##Part 1
Loading and preprocessing the data

```{r loadprocess, echo=TRUE}
# set working directory where the file activity is
#setwd("C:/Users/...")

rawd <- read.csv("activity.csv")

# see dimensions, names of columns and structure of the dataset
dim(rawd)
names(rawd)
str(rawd)

# dates seems to be off, transform from factors to dates

rawd$date <- as.Date(rawd$date)

```

##Part 2
Calculate total steps per day  
Plot histogram  
Report mean and median  

```{r meansteps, echo=TRUE,fig.path ='figure/'}

steps_per_day <- tapply(rawd$steps,rawd$date ,sum)

print(as.table(steps_per_day),type="html")
# note that NAs are ignored here and therefore are not set to zero!

# plot historgram with 10 intervals
hist(steps_per_day,10,col="red",xlab="Steps per day")

# calculate mean and median
mean_steps <- mean(steps_per_day,na.rm=TRUE)
median_steps <- median(steps_per_day,na.rm=TRUE)
```

Mean steps per day is `r mean_steps`.   
Median steps per day is `r median_steps`.  

##Part 3
Plot steps for each 5 minute interval  
Find max steps interval  


```{r interval, echo=TRUE,fig.path ='figure/'}
# get sum of steps per interval for all days
steps_per_interval <- (tapply(rawd$steps,rawd$interval ,mean ,na.rm=TRUE))

# plot pattern

plot(y=steps_per_interval,x=names(steps_per_interval), type="l",
     xlab="5-minute intervals",ylab="steps per interval")
title("Steps pattern per interval - Mean across all days")

# get max
max_interval <- names(which(steps_per_interval==max(steps_per_interval)))
```

 The interval with maximum average steps taken per day is `r max_interval`. i.e. at 8:35 morning.  


##PArt 4A  
missing data handling  

```{r missingdata, echo=TRUE}

na_rows <- rawd

na_rows <- is.na(rawd$steps)
total_na <- sum(na_rows)

# The strategy to fill in the missing data is to use the mean for that 5-minute interval.
# make a copy of raw data

rawd2 <- rawd

for (ii in 1:length(na_rows)) {
        
        if (na_rows[ii]==TRUE) {
                
                ps_interval <- rawd$interval[ii]
                ps_position <- which(names(steps_per_interval)==ps_interval)
                
                rawd2$steps[ii] <- steps_per_interval[ps_position][[1]]
        }
        
}

```

The total number of rows with missing data is `r total_na`.  

The strategy to fill in the missing data is to use the mean for that 5-minute interval.  

The updated dataset is saved as rawd2.

##PArt 4B
histogram  
mean and median  

```{r historaw2, echo=TRUE,fig.path ='figure/'}

steps_per_day2 <- tapply(rawd2$steps,rawd2$date ,sum )

# plot historgram with 10 intervals
hist(steps_per_day2,10,col="blue",xlab="Steps per day")

# let's compare this with previous histogram
par(mfrow=c(1,2))
hist(steps_per_day,10,col="red",ylim=c(0,20),main="Original Data",xlab="Steps per day")
hist(steps_per_day2,10,col="blue",ylim=c(0,20),main="Updated Data",xlab="Steps per day")

# more steps are registered towards the middle of the histogram
# this is expected since the strategy was to fill missing data with the mean.

# calculate mean and median
mean_steps2   <- mean(steps_per_day2)
median_steps2 <- median(steps_per_day2)

```

Mean steps per day with updated data is `r mean_steps2`.  
Median steps per day with updated data is `r median_steps2`.  

Mean before updating data was `r mean_steps`.    
Median before updating data was `r median_steps`.  

Mean hasn't changed at all, since all the NAs were for WHOLE days, therefore the missing values were  
replaced with already calculated mean values from other days. Hence the total mean is unchanged.  

Median has slighly increased leaning towards the mean, but not a big impact. Reason is same as above.  


```{r printnewdata, echo=TRUE}
print(as.table(steps_per_day2),type="html")

# New steps per day can be seen above. The impact is that the sum of total steps is obviously higher now
# Also, you can see that days where there was no steps registered before, are now filled with steps.

```

## PArt 5
Weekdays examination  
weekday vs weekend  


```{r weekdays, echo=TRUE,fig.path ='figure/'}
library(lattice)

# add a new columnn indicating weekend or weekday
rawd2$days <- weekdays(rawd2$date)
rawd2$days <- ifelse(rawd2$days=="Sunday" | rawd2$days=="Saturday", "weekend","weekday")
rawd2$days <- as.factor(rawd2$days)

# Let's see the new variable
head(rawd2)
str(rawd2$days)

# split data into two and calculate mean per interval
weekend <- rawd2[rawd2$days=="weekend",]
weekdays <- rawd2[rawd2$days=="weekday",]

steps_weekend <- (tapply(weekend$steps,weekend$interval ,mean ,na.rm=TRUE))
steps_weekdays <- (tapply(weekdays$steps,weekdays$interval ,mean ,na.rm=TRUE))

# make them data frame
steps_weekend <- data.frame(names(steps_weekend),steps_weekend,rep("weekend",length(steps_weekend)))
steps_weekdays <- data.frame(names(steps_weekdays),steps_weekdays,rep("weekday",length(steps_weekdays)))
names(steps_weekend) <- c("interval","steps","days")
names(steps_weekdays) <- c("interval","steps","days")

# join them
steps_days <- rbind(steps_weekend, steps_weekdays)
steps_days$interval <- as.numeric(as.character(steps_days$interval))

# see part of the result
head(steps_days)

# plot pattern

xyplot(steps ~ interval | days, data = steps_days, type="l", layout = c(1, 2),
       main = "Steps pattern per interval - Mean across all days")


```



